% This must be in the first 5 lines to tell arXiv to use pdfLaTeX, which is strongly recommended.
\pdfoutput=1
% In particular, the hyperref package requires pdfLaTeX in order to break URLs across lines.

\documentclass[11pt]{article}

% Remove the "review" option to generate the final version.
\usepackage{ACL2023}

\usepackage{newtxtext,newtxmath}
\usepackage{latexsym}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{inconsolata}

\usepackage{cleveref}
\usepackage{hyperref}

% Additional packages

% Define \blfootnote command for footnote without marker
\newcommand\blfootnote[1]{%
  \begingroup
  \renewcommand\thefootnote{}\footnote{#1}%
  \addtocounter{footnote}{-1}%
  \endgroup
}

\title{English Nativization Algorithm for Filipino Loan Words}
\setlength\titlebox{10cm}
\author{
  Clarence Ang \\
    De La Salle University \\
    \texttt{clarence\_ang@dlsu.edu.ph} \\\AND%
  Clive Ang \\
    De La Salle University \\
    \texttt{clive\_ang@dlsu.edu.ph} \\\AND%
  Roan Campo \\
    De La Salle University \\
    \texttt{roan\_campo@dlsu.edu.ph} \\\AND%
  Zhean Ganituen \\
    De La Salle University \\
    \texttt{zhean\_robby\_ganituen@dlsu.edu.ph} \\\AND%
  Nathaniel Oco \\
    De La Salle University \\
    \texttt{nathaniel.oco@dlsu.edu.ph} \\
}

\begin{document}
\maketitle
\begin{abstract}
We present a rule-based nativization algorithm for converting English loan words to Filipino orthography. The algorithm applies phonological transformation rules derived from the Komisyon sa Wikang Filipino (KWF) orthography guidelines. We evaluate the method on three datasets: generic drug names, drug brand names, and common English words used in Filipino. The algorithm achieves 88.94\% accuracy on generic drug names, 67.18\% on brand names, and 36\% on common English words. The results indicate that while rule-based approaches are effective for standardized terminology such as drug names, they face limitations when applied to words with irregular pronunciations.
\end{abstract}

\input{chapters/introduction.tex}
\input{chapters/algorithm.tex}
\input{chapters/experiments.tex}
\input{chapters/results.tex}
\input{chapters/conclusion.tex}


\section*{Limitations}

The proposed algorithm has several limitations that should be considered when interpreting the results and applying the method to other contexts.

\paragraph{Grapheme-based approach.} The algorithm operates purely on orthographic representations and does not utilize phonetic information. English orthography lacks a consistent one-to-one correspondence between graphemes and phonemes. For instance, the letter ``i'' in ``diphenhydramine'' represents two distinct sounds: /a\textsc{i}/ (as in ``eye'') for the first occurrence and /\textsc{i}/ (as in ``bit'') for the second. Without access to phonetic transcriptions or pronunciation dictionaries, the algorithm cannot reliably distinguish such cases, leading to incorrect nativizations.

\paragraph{Domain specificity.} The transformation rules were derived primarily from patterns observed in pharmaceutical terminology. Generic drug names follow relatively standardized naming conventions based on International Nonproprietary Names (INN), which exhibit consistent phonological patterns. The lower accuracy on common English words (36\%) indicates that these rules do not transfer well to the broader lexicon, where historical borrowing patterns, regional variations, and irregular pronunciations introduce inconsistencies that rule-based methods cannot capture.

\paragraph{Gold standard construction.} The reference nativizations were created by the authors, who are native Filipino speakers, following the KWF orthography guidelines. However, no inter-annotator agreement was computed, and nativization can be inherently subjective. Different speakers may produce varying nativized forms for the same source word depending on dialect, education, and individual pronunciation habits.

\paragraph{Scope of source languages.} This work addresses only English-to-Filipino nativization. The Filipino language has historically borrowed from multiple languages, including Spanish (e.g., ``kutsara'' from \textit{cuchara}) and Hokkien Chinese (e.g., ``pansit'' from \textit{pi\^{a}n-sit}). The nativization patterns for these source languages differ from English and are not covered by the proposed algorithm.

\section*{Ethics Statement}

The datasets used in this study were compiled from publicly available sources. The generic drug names consist of International Nonproprietary Names (INN), which are standardized designations published by the World Health Organization. The brand names were obtained from the Philippines Food and Drug Administration's public verification database. The common English words were collected from publicly accessible posts on the r/Tagalog subreddit. No private, proprietary, or personally identifiable information was collected or used in this research.

This algorithm is developed for research purposes and is not intended for direct clinical or pharmaceutical application. We acknowledge that nativization of drug names in healthcare settings carries inherent risks. Errors in nativization could result in phonetically similar outputs for distinct drug names, potentially contributing to confusion and medication errors. Therefore, any deployment of this or similar tools in clinical contexts should incorporate verification by qualified healthcare professionals and should not replace established drug name safety protocols.

\section*{Acknowledgments}
This work represents a preliminary work as part of an ongoing undergraduate thesis research project by Zhean Ganituen and others, advised by Nathaniel Oco, on computing confusible drug names in the Philippine setting.

% Entries for the entire Anthology, followed by custom entries
\bibliography{anthology,custom}
\bibliographystyle{acl_natbib}

% \appendix

% \section{Example Appendix}
% \label{sec:appendix}

% This is a section in the appendix.

\end{document}
